Topics in Statistical Theory --- Example Sheet 1

### 2020-10-21 topics_statistical_theory/C1.tex
worked hard on topics :-)

\begin{theorem}[Bennett's inequality]
	Let $X_1, \dotsc, X_n$ be independent random variables with $\EE[X_i] = 0$, $\sigma_i^2 \ceq \Var(X_i) < \infty$, and $X_i \leq b$ for some $b > 0$. Define  $S \ceq \sum_{i=1}^n X_i$, $\nu \ceq \nu$ and $\phi \colon \RR \to \RR$ by $\phi(u) \ceq e^u - 1 - u = \sum_{k=2}^\infty \frac{u^k}{k!}$, then for every $t > 0$ we have
	\[
	\log\EE[e^{tS}] \leq \frac{n \nu }{b^2} \phi(bt). 
	\]

@@

Following this proposition, we can prove Bernstein's inequality:
\begin{theorem}[Bernstein's inequality]
	Let $X_1, \dotsc, X_n$ be independent random variables with $\EE[X] = 0$, $\frac1n\sum_{i=1}^n \Var(X_i) \leq \sigma^2$ and $\frac1n \sum_{i=1}^n \EE[(X_i)_+^q] \leq q! \sigma^2 c^{q-2}/2$ some $\sigma, c > 0$ and for all $q \geq 3$. Then $S \ceq \sum_{i=1}^n X_i$ is sub-Gamma in the right tail with variance factor $n\sigma^2$ and scale parameter $c$. In particular we have
	\[
	\PP(\bar X \geq x) \leq \exp\qty(- \frac{nx^2}{2(\sigma^2 + cx)}). 
	\]
\end{theorem}

>>>>

\begin{theorem}[Bennett's inequality]
	Let $X_1, \dotsc, X_n$ be independent random variables with $\EE[X_i] = 0$, $\sigma_i^2 \ceq \Var(X_i) < \infty$, and $X_i \leq b$ for some $b > 0$. Define  $S \ceq \sum_{i=1}^n X_i$, $\nu \ceq \overline{\Var(X_i)}$ and $\phi \colon \RR \to \RR$ by $\phi(u) \ceq e^u - 1 - u = \sum_{k=2}^\infty \frac{u^k}{k!}$, then for every $t > 0$ we have
	\[
	\log\EE[e^{tS}] \leq \frac{n \nu }{b^2} \phi(bt). 
	\]

@@

Following this proposition, we can prove Bernstein's inequality:
\begin{theorem}[Bernstein's inequality]
	Let $X_1, \dotsc, X_n$ be independent random variables with $\EE[X] = 0$, $\frac1n\sum_{i=1}^n \Var(X_i) \leq \sigma^2$ and $\frac1n \sum_{i=1}^n \EE[(X_i)_+^q] \leq q! \sigma^2 c^{q-2}/2$ some $\sigma, c > 0$ and for all $q \geq 3$. Then $S \ceq \sum_{i=1}^n X_i$ is sub-Gamma in the right tail with variance factor $n\sigma^2$ and scale parameter $c$. In particular we have
	\[
	\PP(\bar X \geq x) \leq \exp\qty(- \frac{nx^2}{2(\sigma^2 + cx)}),
	\]
	for all $x \geq 0$. 
\end{theorem}

### 2020-10-21 topics_statistical_theory/topics_ES1.tex
worked hard on topics :-)

% Functions that appear frequently
\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Span}{Span}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator\Exp{Exp}
\DeclareMathOperator\Beta{Beta}
\DeclareMathOperator\Gam{Gamma}

@@

Finally, define $T = S_{n+1} - S_j$, so that $S_j$ and $T$ are independent. It is known that $S_j \sim \Gam(j, 1)$, $T \sim \gamma(n - j + 1)$, and furthermore that
\[
\frac{S_j}{S_{n+1}} = \frac{S_j}{S_j + T} \deq \frac{\Gamma(j, 1)}{\Gamma(j, 1) + \Gamma(n - j + 1, 1)} \sim \Beta(j, n - j + 1). 
\]
\end{solution}

@@

\begin{proof}
    Define a new distribution $Y = \ind_{X \notin A}$. Its distribution function is given by
    \[
    F_Y(y) = \begin{cases}
        0 &\text{$y < 0$}; \\
        P(A) &\text{$y \in [0, 1)$}; \\
        1 &y \geq 1. 
    \end{cases}, \quad F_Y(0) = P(A). 
    \]
    The empirical distribution function of $Y_1, \dotsc, Y_n \iid Y$ is given by
    \[
    \hat F_n(y) = \frac1n \sum_{i=1}^n \ind_{Y_i \leq y},
    \]
    and thus for $y = 0$ we have
    \[
    \hat F_n(0) = \frac1n \sum_{i=1}^n \ind_{Y_i \leq 0} = \frac1n \sum_{i=1}^n \ind_{U_{i} \in A} = \hat P_n(A).
    \]

    The result now follows from the DKW inequality. \TODO make more precise. 
\end{proof}
\end{document}

>>>>

% Functions that appear frequently
\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Span}{Span}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator\Exp{Exp}
\DeclareMathOperator\Beta{Beta}
\DeclareMathOperator\Gam{Gamma}
\DeclareMathOperator\Cov{Cov}

@@

Finally, define $T_j = S_{n+1} - S_j$, so that $S_j$ and $T_j$ are independent. It is known that $S_j \sim \Gam(j, 1)$, $T_j \sim \gamma(n - j + 1, 1)$, and furthermore that
\[
\frac{S_j}{S_{n+1}} = \frac{S_j}{S_j + T} \deq \frac{\Gamma(j, 1)}{\Gamma(j, 1) + \Gamma(n - j + 1, 1)} \sim \Beta(j, n - j + 1). 
\]
\end{solution}

@@

\begin{proof}
    Define a new distribution $Y = \ind_{X \notin A}$. Its distribution function is given by
    \[
    F_Y(y) = \begin{cases}
        0 &\text{$y < 0$}; \\
        P(A) &\text{$y \in [0, 1)$}; \\
        1 &y \geq 1. 
    \end{cases}
    \]

    The empirical distribution function of $Y_1, \dotsc, Y_n \iid Y$ is given by
    \[
    \hat F_n(y) = \frac1n \sum_{i=1}^n \ind_{Y_i \leq y},
    \]
    and thus for $y \in [0, 1)$ we have
    \[
    \hat F_n(y) = \frac1n \sum_{i=1}^n \ind_{Y_i \leq y} = \frac1n \sum_{i=1}^n \ind_{X_i \in A} = \hat P_n(A).
    \]

    By the DKW inequality we find
    \begin{align*}
    	\PP\qty(\abs{\hat P_n(A) - P(A)} > \eps) = \PP \qty(\sup_{y \in \RR} \abs{\hat F_n(y) - F(y)} > \eps) \leq 2e^{-2n\eps^2}.
    \end{align*}
\end{proof}

\begin{question}
	Let $X \sim \Rm{Bin}(n, p)$. Compare the Hoeffding, Bennett, and Bernstein upper bounds on $\PP(X/n \geq \frac12)$ as $p \to 0$. 
\end{question}

\begin{solution}
	Note that $X/n$ is the average of $n$ i.i.d.\ random variables $Y_i \sim \Rm{Bern}(p)$, where $Y_i \in [0, 1]$ for all $i$.

	We start with Hoeffding's inequality. In this case, we have
	\[
	\PP(X/n - p \geq x) \leq \exp(-\frac{2n^2(1/2)^2}{n}) = \exp(- \frac n2),
	\]
	and this bound also holds as $p \to 0$.

	We continue with Bennett's inequality. We consider the random variables $Y_i - p$, which are bounded from above by $b = 1 - p$. Now Bennett's inequality tells us, with $\sigma_p^2 = \Var(Y_i - p) = p(1-p)$ that
	\begin{align*}
		\PP(X/n \geq x) &\leq \exp(- \frac{n}{(1-p)^2} h\qty(\frac{1-p}{2p(1-p)})) = \exp\qty(- \frac{n}{(1-p)^2} h\qty(\frac{1}{2p})).
	\end{align*}
We compute
\[
h\qty(\frac1{2p}) = (1 + \frac1{2p}) \log(1 + \frac1{2p}) - \frac1{2p} = \log(1 + \frac1{2p}) + \frac1{2p} \qty(\log(1 + \frac1{2p}) - 1) \overset{p \downarrow 0}\to +\infty. 
\]
Since $\frac{n}{(1-p)^2}$ is clearly bounded by $n$, we conclude that 
\[
\PP(X/n \geq x) \to e^{-\infty} = 0. 
\]

%   Bennett's inequality tells us that
%	\[
%	\PP(X/n \geq \frac12) \leq \exp\qty(- n\nu h(1/(2\nu))), 
%	\]
%	where $h(u) = (1 + u) \log(1+ u)$ and $\nu = p(1-p)$. Plugging this in yields
%	\[
%	\exp(-n\nu h(x/\nu)) = \exp(-n p(1-p) \qty(1 + \frac{1}{2p(1-p)})\log(1 + \frac{1}{2p(1-p)}))
%	\]

We finish with Bernstein's inequality.  We have for $q \geq 3$ and $p \leq\frac12$ that 
\begin{align*}
	\EE[(Y_i - p)_+^q] &= p(1-p)^q = \sigma_p^2 (1-p)^{q-1} = (q! \sigma_p^2 (1-p)^{q-2}/2)\cdot (2(1-p)/q!) \\
	&\leq q!\sigma_p^2 ((1-p)/3)^{q-2}/2.
\end{align*}

Now Bernstein's inequality tells us that
\[
\PP(X/n \geq \frac12 \leq \exp\qty(- \frac{n(1/2)^2}{2(\sigma_p^2 + (1-p)/6)}) = \exp(- \frac{n}{8\sigma_p^2 + 4(1-p)/3}) \overset{p \to 0}{\to} \exp(-\frac{3n}{4}). \]

Of course, the true limit is 0 for any $n$, which is only given by Bennett's inequality. We also see that Hoeffding's inequality gives the worst result. 
\end{solution}

\begin{question}
	Derive the following alternative form of Bernstein's inequality: under the same conditions, 
	\[
	\PP\qty(\bar X \geq \sqrt{\frac{2\sigma^2 \log(1/\delta)}{n}} + \frac cn \log(1/\delta)) \leq \delta
	\]
	for every $\delta \in (0, 1]$. 
\end{question}

\begin{proof}
	We have 
	\[
	\PP(\bar X \geq x) \leq \exp(- \frac{nx^2}{2(\sigma^2 + cx)}) \eqqcolon \delta. 
	\]
	Now we just need express $x$ in terms of $\delta$: taking logarithms on both sides we obtain
	\[
	- \frac{nx^2}{2(\sigma^2 + cx)} = \log(\delta) \implies nx^2 = 2(\sigma^2 + cx) \log(1/\delta) \implies nx^2 - 2c\log(1/\delta) x - 2\sigma^2\log(1/\delta) = 0.
	\]

	Using the abc-formula with the fact that $x \geq 0$ yields 
	\begin{align*}
	x &= \frac{2c \log(1/\delta) + \sqrt{4c^2\log^2(1/\delta) + 8n\sigma^2\log(1/\delta)}}{2n} \\
	&= \frac cn \log(1/\delta) + \sqrt{\frac{c^2}{n^2}\log^2(1/\delta) + \frac{2\sigma^2}{n} \log(1/\delta)} \\
	&= ???? 
	\end{align*}
\end{proof}

\begin{question}
	\begin{enumerate}[(a)]
		\item Let $X_1, \dotsc, X_n \iid F$ and let $\hat F_n$ denote their empirical distribution function. For $t_1 < \dotsb < t_k$, write down the distribution of 
		\[
		n\qty(\hat F_n(t_1), \hat F_n(t_2) - \hat F_n(t_1), \dotsc, \hat F_n(t_k) - \hat F_n(t_{k-1}), 1 - \hat F_n(t_k)). 
		\]
		\item Find the asymptotic distribution of $n^{1/2} (\hat F_n(t_1) - F(t_1), \dotsc, \hat F_n(t_k) - F(t_k))$. 
	\end{enumerate}
\end{question}

\begin{solution}
	\begin{enumerate}[(a)] \item
	Write $n \hat F_n(t) = \sum_{i=1}^n \ind_{X_i \leq t} = \# \qty{i \mid X_i \leq t}$, and analogously, for $t < u$, 
	$n (\hat F_n(u) - \hat F_n(t)) = \#\qty{i \mid t < X_i \leq u}$.

	Then, defining $t_0 = -\infty$ and $t_{k+1} = \infty$,  we find that
	\begin{gather*}
	\PP\qty[ n \qty(\hat F_n(t_1), \dotsc, 1 - \hat F_n(t_k)) (a_1, \dotsc, a_{n+1})] \\ = \PP[\text{exactly $a_i$ of the $X_i$ lie in $(t_{i-1}, t_i]$ for $i = 1, \dotsc, n$}].
	\end{gather*}

This probability is 0 unless all $a_i$ are nonnegative integers with $\sum_i a_i = n$. In this case, we can first choose $a_1$ of the $X_i$ which must lie in $(\infty, t_1]$, then we choose $a_2$ of the remaining $X_i$ which must lie in $(t_1, t_2]$ and so forth. We find that the above probability equals
\[
\binom{n}{a_1} (F(t_1))^{a_1} \binom{n - a_1}{a_2} (F(t_2) - F(t_1))^{a_2} \dotsb \binom{a_k + a_{k_1}}{a_k} (F(t_k) - F(t_{k-1}))^{a_k} \cdot (1 - F(t_k))^{a_{k+1}}. 
\]

\item By the central limit theorem, the asymptotic distribution is $N(0, \Sigma)$, where $\Sigma$ is the covariance matrix of $(\hat F_n(t_1), \dotsc, \hat F_n(t_k))$. We will compute the entries of $\Sigma$.

Choose $t \in \RR$ arbitrarily. Then we have
\begin{align*}
	\Var(\hat F_n(t)) &= \EE[\hat F_n^2(t)] - \EE[\hat F_n(t)]^2 = \EE\qty[\qty(\frac1n\sum_i \ind_{X_i \leq t})^2] - F^2(t) \\
	&= \frac1{n^2}\EE\qty[\sum_i \ind_{X_i \leq t} + 2\sum_{i < j} \ind_{X_i \leq t} \ind_{X_j \leq t}] - F^2(t) \\
	&= \frac{F(t) + (n-1) F^2(t)}{n} - F^2(t) = \frac{F(t)(1 - F(t))}{n},
\end{align*}
so we have computed the diagonal entries $\Sigma_{ii} = \frac{F(t_i)(1 - F(t_i))}{n}$.

Now we must compute the covariances: assume $s < t$, then  
\begin{align*}
	\Cov(\hat F_n(s), \hat F_n(t)) &= \EE[\hat F_n(s) \hat F_n(t)] - \EE[\hat F_n(s)] \EE[\hat F_n(t)] \\
	&= \frac1{n^2} \sum_{i, j} \EE[\ind_{X_i \leq s} \ind_{X_j \leq t}] - F(s)F(t) \\
	&= \frac1{n^2} \qty(n F(s) + n(n-1) F(s)F(t)) - F(s)F(t) \\
	&= \frac{F(s) + (n-1) F(s) F(t)}{n} - F(s)F(t) = \frac{F(s) - F(s)F(t)}{n}. 
\end{align*}
This gives the diagonal entries $\Sigma_{ij} = \frac{F(t_i) - F(t_i)F(t_j)}{n}$ for $i < j$. 
In the end, we find
\[
\Sigma_{ij} = \frac1n \cdot \begin{cases}
	F(t_i)(1 - F(t_i)) &\text{if $i = j$}, \\
	F(t_{\min(i, j)}) - F(t_i) F(t_j) &\text{if $i \neq j$}. 
\end{cases}
\]
\end{enumerate} 
\end{solution}
\end{document}
